---
title: "TaxData Enhancement Project"
author: Don Boyd
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_notebook:
    toc: true
    toc_depth: 5
    df_print: paged

---

# Project Statement of Work

> Augment Boyd’s current project in which he is developing methods for creating a single state microdata tax file, in two ways: (a) rigorously compare file quality under several different approaches under investigation, and (b) port the approach that produces the best results from R to Python so that it can readily be integrated with TaxData.


<!-- First set things up, then run things -->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


```{r system_specific_info, include=TRUE}
# change these as needed
pufdir <- "C:/Users/donbo/Dropbox/OSPC - Shared/IRS_pubuse_2011/"

# Note that my .Rprofile file runs these commands:
# RPROJ <- list(PROJHOME = normalizePath(getwd()))
# attach(RPROJ)
# rm(RPROJ)

# It does this because rmd files change the home directory.
# This creates a PROJHOME variable that points to the project main directory that can be used when saving and reading files.
# See how it is used in file reads below.


```


```{r libraries, message=FALSE, warning=FALSE, include=FALSE}
library("magrittr")
library("plyr") # needed for ldply; must be loaded BEFORE dplyr
library("tidyverse")
options(tibble.print_max = 60, tibble.print_min = 60) # if more than 60 rows, print 60 - enough for states
# ggplot2 tibble tidyr readr purrr dplyr stringr forcats

library("scales")
library("hms") # hms, for times
library("lubridate") # lubridate, for date/times
library("vctrs")

library("grDevices")
library("knitr")
library("kableExtra")

# library("zoo") # for rollapply

# library("forecast")

library("ipoptr")

```



```{r general_functions, message=FALSE, warning=FALSE, include=FALSE}
ns <- function(df) {names(df) %>% sort} # name sort

ht <- function(df, nrecs=6){
  print(utils::head(df, nrecs))
  print(utils::tail(df, nrecs))
}

```


# Make sure that `ipoptr` works by solving the simple "banana" test problem included with the package
Solve the Rosenbrock Banana test problem, a nonlinear function of 2 variables.

Don't worry now about reading the output. If you see "EXIT: Optimal Solution Found." then everything worked fine.

```{r echo=TRUE, message=FALSE, warning=FALSE}
## Rosenbrock Banana function
eval_f <- function(x) {   
  return( 100 * (x[2] - x[1] * x[1])^2 + (1 - x[1])^2 )
}

## Gradient of Rosenbrock Banana function
eval_grad_f <- function(x) { 
  return( c( -400 * x[1] * (x[2] - x[1] * x[1]) - 2 * (1 - x[1]),
             200 * (x[2] - x[1] * x[1])) )
}

# The Hessian for this problem is actually dense, 
# This is a symmetric matrix, fill the lower left triangle only.
eval_h_structure <- list( c(1), c(1,2) )

eval_h <- function( x, obj_factor, hessian_lambda ) {
  return( obj_factor*c( 2 - 400*(x[2] - x[1]^2) + 800*x[1]^2,      # 1,1
                        -400*x[1],                                 # 2,1
                        200 ) )                                    # 2,2
}

# initial values
x0 <- c( -1.2, 1 )

opts <- list("print_level" = 5,
             "file_print_level" = 12,
             "output_file" = "banana.out",
             "tol"=1.0e-8)

# solve Rosenbrock Banana function with analytic hessian 
print(ipoptr( x0=x0, 
               eval_f=eval_f, 
               eval_grad_f=eval_grad_f, 
               eval_h=eval_h,
               eval_h_structure=eval_h_structure,
               opts=opts) )

```


# Set things up so that we can test `ipoptr` on PUF targeting problems

## Create a PUF subset to work with
ONETIME: First, create a small PUF test data set from the PUF that we will work with. See chunk code for information.

```{r message=FALSE, warning=FALSE, include=FALSE, eval=FALSE}
# You will have to run this one time on your own machine - set include=TRUE
# Note that I have the data directory in my .gitignore file so that the data - which are restricted in use - are not uploaded to github
pufraw <- read_csv(paste0(pufdir, "puf2011.csv"), col_types = cols(.default= col_double()))
# glimpse(pufraw)
ns(pufraw)

# keep a few variables:
# RECID unique id
# E00100 AGI
# E00200 wages
# E00300 interest received -- not kept
# E01700 pension income in AGI
# E02000 schedule E net income 
# S006 record weight as an integer - must be divided by 100
vars <- c("RECID", "E00100", "E00200", "E01700", "E02000", "S006")
set.seed(1234)
pufsub <- pufraw %>% 
  sample_n(5000) %>%
  select(vars) %>%
  mutate(weight=S006 / 100, otheragi=E00100 - E00200 - E01700 - E02000) %>%
  select(-S006) %>%
  select(RECID, weight, everything())

# glimpse(pufsub)
# ht(pufsub)
saveRDS(pufsub, paste0(PROJHOME, "/data/pufsub.rds")) # make sure data folder is in .gitignore

```


Get the PUF subset data and show its contents. We kept 5,000 records, selected randomly, with the following information:

+ RECID unique id
+ E00100 AGI
+ E00200 wages
+ E01700 pension income in AGI
+ E02000 schedule E net income (Rental real estate, royalties, partnerships, S corporations, trusts, etc.)
+ otheragi calculated as E00100 minus the other items
+ weight (the original S006 record weight, which we divided by 100 so that it is in the correct units)

```{r}
pufsub <- readRDS(paste0(PROJHOME, "/data/pufsub.rds")) 
# glimpse(pufsub) - do not show individual records - they are restricted and the html file is uploaded to github
summary(pufsub)
```



## Define an objective function for use with PUF targeting
### What does an objective function look like?
We're going to choose new weights for the PUF that will achieve some targets (constraints). In these exaples, we want to choose weights that are close to the original weights on the file, but hit the targets. We'll penalize new weights that are far from the old weights.

We'll compute the ratio of the new weight to the current weight - call it x (a vector of weights, one for each record) and minimize the following objective (penalty) function based upon this ratio:

$$min \sum_{x} w_1(x^2 + x^{-2} -2)$$

where w1 is the original weight and x is the ratio of the new weight (call it w2) to the original weight (i.e., $x=w_2 / w_1$).

For any single record, the penalty function is minimized if the ratio, x, is 1. It penalizes proportionate differences from 1 equally. That is, if the weight is reduced to 10% of the original weight (x=.1) the penalty is the same as if the weight is increased to 10 times the original weight (x=10). Far distances from 1 are penalized disproportionately relative to smaller distances.

Because the penalty is weighted by the current weight, a change to the weight on a record that is very important (has a large weight) will have a greater penalty than a change to the weight on a record that is less important.

The minimum possible value for the function, summed over all x's, is zero of course.

The minus 2 is not necessary analytically but it makes the objective function easier to interpret and also reduces risk of numerical instability by keeping the objective function calculation for each weight close to zero.

The following graph shows the objective function for x ranging from 0.1 to 10, where w1 is assumed away (weight is 1 for all records). It is minimized at 1, of course, and the values at 0.1 and 10.0 are the same.

```{r}
# graph of the objective  assuming all weights are 1.
f1 <- function(x) x^2 + x^-2 - 2
 
tibble(x=seq(0.01, 100, .01)) %>%
  mutate(y=f1(x)) %>%
  filter(x > 0.09, x < 10.01) %>%
  ggplot(aes(x, y)) +
  geom_line() +
  geom_vline(xintercept = 1, linetype="dashed") +
  scale_x_continuous(name="x (ratio of new weight to old weight)", 
                     breaks=c(.1, .25, .5, .75, 1, 2:20)) +
  theme(axis.text.x = element_text(angle = -90, vjust=0.5)) +
  scale_y_continuous(name="y (penalty)") +
  ggtitle("Objective function y=x^2 + x^-2 - 2, over the interval [0.1, 10]")


```

This is not the only possible objective function. Two other candidates are:

+ The TaxData objective function: minimize the sum of the absolute value of the percentage change from the original weight to the new weight (or, identically, minimize the sum of the difference between x and 1), or $min \sum_x abs(x - 1)$
+ An alternative that has been put forth $min \sum_x w_1(x - 1)^2$

Note that the function shown in the previous graph is symmetric around 1 in x and its multiplicative inverse, while both of these are additively symmetric around 1.

Note also that the TaxData function is not continuous or differentiable. Also, it is not weighted by the current weight.

```{r}

f2 <- function(x) abs(x - 1) # the TaxData function
f3 <- function(x) (x - 1)^2 # a continuous differentiable variant of the TaxData function

tibble(x=seq(0.01, 100, .01)) %>%
  filter(x > 0.09, x < 10.01) %>%
  mutate(y1=f1(x), y2=f2(x), y3=f3(x),
         # scale the functions to have the same value at x==10
         y2=y2 * y1[x==10] / y2[x==10],
         y3=y3 * y1[x==10] / y3[x==10]) %>%
  gather(objfn, y, y1, y2, y3) %>%
  mutate(objfn=factor(objfn, 
                      labels=c("y1: x^2 + x^-2 - 2",
                               "y2: abs(x - 1)",
                               "y3: (x - 1)^2"))) %>%
  ggplot(aes(x, y, colour=objfn)) +
  geom_line() +
  geom_vline(xintercept = 1, linetype="dashed") +
  scale_x_continuous(name="x (ratio of new weight to old weight)", 
                     breaks=c(.1, .25, .5, .75, 1, 2:20)) +
  theme(axis.text.x = element_text(angle = -90, vjust=0.5)) +
  scale_y_continuous(name="y (penalty)\nscaled so that all have the same penalty at x==10") +
  ggtitle("Selected objective functions over the interval [0.1, 10]")


```


### Implement an objective function and its gradient
Define R functions that will be used within ipoptr (and can be used externally, too) to calculate, for any given x vector:

+ The objective function evaluated at point x (a vector). It returns a single value.
+ The gradient of the objective function evaluated at point x. It returns a vector of values, the same length as x, where each element is the partial derivative of the objective function with respect to the corresponding element in x. This is used by IPOPT in choosing a search direction. While some non-linear programming (NLP) methods do not require derivatives, IPOPT does. While it is possible to write a function that calculates approximate derivatives, it is better to specify analytic derivatives when possible.

In addition to passing x to each function we will pass a list, that I call `inputs`, that can have anything we want it to have. The reason for this is that `ipoptr` requires any function it uses to receive the same arguments. Passing the same list to each function allows us to have functions that use different items (all contained within inputs) but receive the same arguments.

We could call these functions anything. I call the objective function `eval_f_xtop` which uses the name `ipoptr` uses for the objective (eval_f) plus a suffix, xtop, which stands for x to the power p (we could use any even power for the objective funciton, not just 2 as in the example above). Similarly, I call the gradient `eval_grad_f_xtop`.


```{r functions_obj, echo=TRUE}
eval_f_xtop <- function(x, inputs) {
  #.. objective function - evaluates to a single number ----
  
  # ipoptr requires that ALL functions receive the same arguments, so a list called inputs is passed to ALL functions
  
  # here are the objective function, the 1st deriv, and the 2nd deriv
  # http://www.derivative-calculator.net/
  # w{x^p + x^(-p) - 2}                                 objective function
  # w{px^(p-1) - px^(-p-1)}                             first deriv
  # p*w*x^(-p-2)*((p-1)*x^(2*p)+p+1)                    second deriv
  
  # make it easier to read:
  p <- inputs$p
  w <- inputs$weight
  
  obj <- sum(w * {x^p + x^(-p) -2})
  
  return(obj)
}


eval_grad_f_xtop <- function(x, inputs){
  #.. gradient of objective function - a vector length x ----
  # giving the partial derivatives of obj wrt each x[i]
  
  # ipoptr requires that ALL functions receive the same arguments, so a list called inputs is passed to ALL functions
  
  # here are the objective function, the 1st deriv, and the 2nd deriv
  # http://www.derivative-calculator.net/
  # w{x^p + x^(-p) - 2}                                 objective function
  # w{px^(p-1) - px^(-p-1)}                             first deriv
  # p*w*x^(-p-2)*((p-1)*x^(2*p)+p+1)                    second deriv
  
  # make it easier to read:
  p <- inputs$p
  w <- inputs$weight
  
  gradf <- w * (p * x^(p-1) - p * x^(-p-1))
  
  return(gradf)
}
```

Calculate the objective function and its gradient for a few values, as a test:

```{r echo=TRUE}
x0 <- c(1, .1, 10)
inputs <- list()
inputs$p <- 2

inputs$weight <- 1
for(i in 1:length(x0)) {print(eval_f_xtop(x0[i], inputs))}

inputs$weight <- c(1, 1, 1)
eval_f_xtop(x0, inputs)
eval_grad_f_xtop(x0, inputs)

```



## Define constraints and associated functions
We will define a set of targets for the reweighted PUF, or "constraints" in NLP nomenclature.

Examples of constraints are:

+ The sum of weighted adjusted gross income (E00100) or another variable in the file using a given set of weights. For example:
++ sum(E00100 * weight) gives the sum with current weights, and
++ sum(E00100 * weight * x) gives the sum with new weights, if x is the ratio of new weights to original weights
+ The number of weighted tax returns in the file
+ The sum of weighted negative values for a variable. For example, E02000 (rental income etc.) can be negative (a loss). We might want to target the sum of these negative values separately from the sum of positive values for the variable.
+ The weighted number of negative values for a variable (e.g., the number of returns, weighted, that have losses).
+ Any of the above, within specific segments of the file, such as the number returns with negative rental income, within the subset of the file that has married couples with income between $50k and $75k. Obviously it only makes sense to construct a target for such a narrow portion of the file if we think we can estimate or forecast what a proper target should be.

The file with existing weights has total AGI of `r scales::comma(sum(pufsub$E00100 * pufsub$weight))` and total wages of `r scales::comma(sum(pufsub$E00200 * pufsub$weight))`. We might (if we had information that suggested this), set a target for AGI that is 10% greater (`r scales::comma(sum(pufsub$E00100 * pufsub$weight)*1.1)`) while at the same time setting a target for wages that is 5% less (`r scales::comma(sum(pufsub$E00200 * pufsub$weight)*.95)`). 


### Creating a set of constraint coefficients using dense matrices
We need to tell `ipoptr` (which tells IPOPT) how the targeted values on our reweighted file will change as we change our variable x -- that is we need to define the first partial derivative of each constraint with respect to each x variable. In our problem, these are constants, so I call them constraint coefficients.

There is more than one way to set up the constraints for the problem. For now, not worrying about computer resources, we'll use dense matrices. Imagine a matrix where every row is a record in our data file, every column represents a possible constraint of interest (e..g, the weighted sum of AGI, or the weighted number of returns with negative rental income), and every cell in the matrix represents how much the constraint (e.g,. the weighted sum of AGI) will change if we change the ratio x applicable to that return.

Suppose that the j-th column of this matrix represents the constraint for the weighted sum of positive AGI; let's name it E00100_sumpos.
Suppose that the i-th row of this matrix, tax return i, has AGI of $10,000 and an original weight of 17. Then cell [i, j] in this matrix would be 10,000 x 17 or 170000 ($170k), which is how much aggregate AGi would increase if we increased x[i] by 1. For example, if we increased the weight for return i by 10% (x[i] is 1.1 and the change in x[i] is 0.1), total weighted AGI would increase by 10,000 x 17 x 10%.


#### Define functions to help in developing constraint coefficients
The functions below calculate constraint coefficients for common constraints of interest. var is a vector of values for a variable -- for example, var[i] in the example above is 10,000 and weight[i] is 17

```{r functions_concoef, echo=TRUE}
# functions to compute constraint coefficients
nnz <- function(var, weight) {(var != 0) * weight} # weighted number of returns with nonzero values
nz <- function(var, weight) {(var == 0) * weight} #  weighted number of returns with zero values
npos <- function(var, weight) {(var > 0) * weight} # weighted number of returns with positive values
nneg <- function(var, weight) {(var < 0) * weight} # weighted number of returns with negative values

sumval <- function(var, weight) {weight * var} # weighted sum of nonzero value
sumpos <- function(var, weight) {(var > 0) * weight * var}
sumneg <- function(var, weight) {(var < 0) * weight * var}

# sum(pufsub$E00100 * pufsub$weight) / 1e9

```


#### Create a constraint-coefficient matrix-like data frame
Now we apply these functions to our data to construct constraint coefficients for potential constraints, for a few selected variables. This is not necessarily the most efficient way to do this. For example we might want to first decide which constraints we care about, and only compute them. With small datasets, this is not an important consideration.

Each column will represent a constraint, and each row will correspond to a row in the data.

```{r echo=TRUE}
ccvars <- c("E00100", "E00200", "E02000")
ccoef_all <- pufsub %>%
  mutate_at(vars(ccvars), 
            list(npos = ~npos(., weight),
                 sumval = ~sumval(., weight),
                 nneg = ~nneg(., weight)))
# glimpse(ccoef_all)
ht(ccoef_all)

```


### Define generic function to calculate constraint sums based on a given data file and an x vector
```{r echo=TRUE}
calc_constraints <- function(ccoef, constraint_vars, x=rep(1, nrow(ccoef))){
  # calculated weighted sums of constraint_vars that are in data, using an optional multiplier x
  # weights: numeric vector
  # uwccoef: df with 1 column per constraint variable (plus possibly additional columns); 1 row per person (tax return), 
  #        values are constraint unweighted coefficients
  # contraint_vars: character vector
  # return: a named vector
  colSums(x * select(ccoef, constraint_vars))
}

```


Check: Print sums on file as it exists, and sums with alternative weights
```{r}
# calculate values on the file and for some randomly chosen targets
vars <- str_subset(names(ccoef_all), "_")
filesums <- calc_constraints(ccoef_all, vars) # a named vector
# check:
# sum(pufsub$weight * pufsub$E00100)
# sum(ccoef_all$E00100_sumval)

# pick some random x's
set.seed(5)
x <- rnorm(nrow(pufsub), mean=1, sd=.2)
targets <- calc_constraints(ccoef_all, vars, x)

cbind(filesums, targets, gapratio=targets/filesums)
```


### Define constraint functions for use within `ipoptr`

`ipoptr` requires:
- A constraint calculation function that, for a given vector x, returns a vector with one value for each constraint calculated using that x. It is based on `calc_constraints` above but also accepts `inputs` as an argument and removes names from the result vector. Internally, ipoptr calls this function `eval_g`; I name our version `eval_g_dense` because it uses dense matrices (or data frames), rather than sparse ones.
- The Jacobian of the constraints - the first partial derivatives of each constraint with respect to each element of the vector x. We use a dense structure that is a vector with 1 element for every partial derivative. In a problem with 5,000 records and 10 constraints the return vector will have 50,000 elements. In our problem (but not all NLP problems) these derivatives are constant and may need to be computed in each iteration of the solver. We put the constant vector into the inputs list and return it as the result of our Jacobian function, unchanged. ipoptr calls this function `eval_jac_g` internally, and I have named our version `eval_jac_g_dense`. It is dense in the sense that many partial derivatives may be zero but we store them and pass them around anyway. In very large problems it can make sense to use sparse matrices that store only the nonzero elements, and keep track of which ones are nonzero.

We also need to tell `ipoptr` the structure of the Jacobian - which values mean what. We will need to pass it a list that has one element for each row of the data. Each element of the list is a vector. The vector has one element for each constraint that indicates the column that the constraint is in. (This is different when we use sparse matrices.) The helper function below, `define_jac_g_structure_dense`, creates the list that `ipoptr` wants.


```{r functions_ipopt, echo=TRUE}

eval_g_dense <- function(x, inputs){
  #.. constraints that must hold in the solution ----
  # just give the LHS of the expression
  # return a vector where each element evaluates a constraint (i.e., sum of (x * a cc matrix column), for each column)
  
  # ipoptr requires that ALL functions receive the same arguments, so a list called inputs is passed to ALL functions
  
  unname(calc_constraints(inputs$ccoef, inputs$constraint_vars, x)) # ipoptr doesn't seem to like a named vector
}


eval_jac_g_dense <- function(x, inputs){
  # Jacobian of the constraints ----
  # the Jacobian is the matrix of first partial derivatives of constraints (these derivatives may be constants)
  # this function evaluates the Jacobian at point x
  
  # This is the dense version that returns a vector with one element for EVERY item in the Jacobian (including zeroes)
  # Thus the vector has n_constraints * n_variables elements
  # first, all of the elements for constraint 1, then all for constraint 2, etc...
  
  # because constraints in this problem are linear, the derivatives are all constants
  
  # ipoptr requires that ALL functions receive the same arguments, so the inputs list is passed to ALL functions
  
  return(inputs$cc_dense)
}


define_jac_g_structure_dense <- function(n_constraints, n_variables){
  #.. function to define the structure of the Jacobian ---
  # list with n_constraints elements
  # each is 1:n_variables
  lapply(1:n_constraints, function(n_constraints, n_variables) 1:n_variables, n_variables)
} 

# Examples of how to use define_jac_g_structure_dense:
# eval_jac_g_structure_dense <- define_jac_g_structure_dense(n_constraints=2, n_variables=4)

# eval_jac_g_structure_dense <- define_jac_g_structure_dense(
#   n_constraints=ncol(data[, names(constraints)]), 
#   n_variables=nrow(data[, names(constraints)]))


```


### Define Hessian function for use within `ipoptr`

`ipoptr` does not strictly require a Hessian matrix of 2nd derivatives but in this problem it seems to work best with one.

The function below computes the Hessian at point x.
```{r hessian, echo=TRUE}
eval_h_xtop <- function(x, obj_factor, hessian_lambda, inputs){
  # The Hessian matrix ----
  # The Hessian matrix has many zero elements and so we set it up as a sparse matrix
  # We only keep the (potentially) non-zero values that run along the diagonal.
  
  # ignore obj_factor and hessian_lambda - they are required arguments but we do not create them.
  
  # http://www.derivative-calculator.net/
  # w{x^p + x^(-p) - 2}                                 objective function
  # w{px^(p-1) - px^(-p-1)}                             first deriv
  # p*w*x^(-p-2)*((p-1)*x^(2*p)+p+1)                    second deriv
  
  # make it easier to read:
  p <- inputs$p
  w <- inputs$weight
  
  hess <- obj_factor * 
    (  p*w*x^(-p-2) * {(p-1)*x^(2*p) + p + 1}  )
  
  return(hess)
}

```


# Start with a simple PUF targeting problem
Later we will move to more-complex problems.

Here we look at a subset of the constraints we computed above. We compute current file totals and make up some perturbed totals that we will use as targets. Then we solve the problem.


## Define the targets
Remind ourselves of the constraint coefficients available: `r names(ccoef_all) %>% sort`
Let's target:

+ The number of returns -- E00100_npos will do that
+ Sum of AGI -- E00100_sumval
+ Sum of wages -- E00200_sumval
+ number of returns that have negative rental income -- E02000_nneg


```{r targets1, message=FALSE, warning=FALSE}
# glimpse(ccoef_all)
# Let's target the number of returns (E00100_npos) will do that
# define constraint variables and their targets
# let's just use the sums

constraint_vars <- c("E00100_npos", "E00100_sumval", "E00200_sumval", "E02000_nneg")

ccoef <- ccoef_all %>%
  select(constraint_vars)
  
# Get the file sums so we know where we are starting from
(filesums <- calc_constraints(ccoef, constraint_vars))
# (filesums <- calc_constraints(ccoef, constraint_vars, x=1))

# create targets with perturbed values
targ_factors <- list(E00100_npos=1.25,
                     E00100_sumval=0.75,
                     E00200_sumval=.9,
                     E02000_nneg=.8) %>% 
  unlist # define multipliers for some values, as a named vector

targets <- filesums[names(targ_factors)] * targ_factors

# Look at our targets
cbind(init_vals=filesums[names(targets)], targets, ratio=targets/filesums[names(targets)])

```


## Solve the problem with no bounds on the x values
```{r optimize1_nobounds, echo=TRUE}
# glimpse(pufsub)
# glimpse(ccoef)
# targets
inputs <- list()
inputs$p <- 2
inputs$weights <- pufsub$weight
inputs$ccoef <- ccoef
inputs$constraint_vars <- constraint_vars
inputs$cc_dense <- c(as.matrix(inputs$ccoef[, inputs$constraint_vars])) # flattens the cc matrix

# create vectors with constrain lower bounds and upper bounds
clb <- targets[inputs$constraint_vars]
cub <- targets[inputs$constraint_vars]

cbind(init_vals=filesums[names(targets)], targets, clb, cub)

eval_jac_g_structure_dense <- define_jac_g_structure_dense(n_constraints=length(inputs$constraint_vars), 
                                                           n_variables=nrow(inputs$ccoef))
# str(eval_jac_g_structure_dense)

eval_h_structure <- lapply(1:length(inputs$weight), function(x) x) # diagonal elements of our Hessian

opts <- list("print_level" = 0,
             "file_print_level" = 5, # integer
             "linear_solver" = "ma27", # mumps pardiso ma27 ma57 ma77 ma86 ma97
             "max_iter"=200,
             "output_file" = "prob1_nobounds.out")
  
result1 <- ipoptr(x0 = rep(1, length(inputs$weights)),
                 eval_f = eval_f_xtop, # arguments: x, inputs
                 eval_grad_f = eval_grad_f_xtop,
                 eval_g = eval_g_dense, # constraints LHS - a vector of values
                 eval_jac_g = eval_jac_g_dense,
                 eval_jac_g_structure = eval_jac_g_structure_dense,
                 eval_h = eval_h_xtop, # the hessian is essential for this problem
                 eval_h_structure = eval_h_structure,
                 constraint_lb = clb,
                 constraint_ub = cub,
                 opts = opts,
                 inputs = inputs)

names(result1) %>% sort
result1$status; result1$message; result1$iterations
result1$objective
quantile(result1$solution, probs=c(0, .05, .1, .25, .5, .75, .9, .95, 1))

calcsums <- calc_constraints(inputs$ccoef, inputs$constraint_vars, x=result1$solution)
cbind(filesums, targets, clb, calcsums, cub)

```


## Rerun with bounds on the x values
```{r optimize1_bounds, echo=TRUE}
# glimpse(pufsub)
# glimpse(ccoef)
targets

inputs <- list()
inputs$p <- 2
inputs$weights <- pufsub$weight
inputs$ccoef <- ccoef
inputs$constraint_vars <- constraint_vars
inputs$cc_dense <- c(as.matrix(inputs$ccoef[, inputs$constraint_vars])) # flattens the cc matrix

# create vectors with constrain lower bounds and upper bounds
clb <- targets[inputs$constraint_vars]
cub <- targets[inputs$constraint_vars]

cbind(init_vals=filesums[names(targets)], targets, clb, cub)

eval_jac_g_structure_dense <- define_jac_g_structure_dense(n_constraints=length(inputs$constraint_vars), 
                                                           n_variables=nrow(inputs$ccoef))
# str(eval_jac_g_structure_dense)

eval_h_structure <- lapply(1:length(inputs$weight), function(x) x) # diagonal elements of our Hessian


# set arbitrary bounds for x that fall inside the solution above
xlb <- rep(.3, length(inputs$weights))
xub <- rep(10, length(inputs$weights))

opts <- list("print_level" = 0,
             "file_print_level" = 5, # integer
             "linear_solver" = "ma27", # mumps pardiso ma27 ma57 ma77 ma86 ma97
             "max_iter"=200,
             "output_file" = "prob1_bounds.out")
  
result1b <- ipoptr(x0 = rep(1, length(inputs$weights)),
                 lb = xlb,
                 ub = xub,
                 eval_f = eval_f_xtop, # arguments: x, inputs
                 eval_grad_f = eval_grad_f_xtop,
                 eval_g = eval_g_dense, # constraints LHS - a vector of values
                 eval_jac_g = eval_jac_g_dense,
                 eval_jac_g_structure = eval_jac_g_structure_dense,
                 eval_h = eval_h_xtop, # the hessian is essential for this problem
                 eval_h_structure = eval_h_structure,
                 constraint_lb = clb,
                 constraint_ub = cub,
                 opts = opts,
                 inputs = inputs)

# names(result1b) %>% sort
paste0("Status: ", result1b$status)
result1b$message
paste0("Iterations: ", result1b$iterations)
result1b$objective; result1$objective
quantile(result1b$solution, probs=c(0, .05, .1, .25, .5, .75, .9, .95, 1))

calcsums <- calc_constraints(inputs$ccoef, inputs$constraint_vars, x=result1b$solution)
cbind(filesums, targets, clb, calcsums, cub)

```


## Add targets for a portion of the distribution of wages
This is not necessarily the most efficient way to do this, but it should be pretty clear.
```{r optimize2_wages, echo=TRUE}

# add constraint coefficients for wages
glimpse(ccoef_all)
wagevals <- ccoef_all %>%
  select(RECID, weight, E00200) %>%
  mutate(wagegroup=cut(E00200, c(-Inf, 0, 10e3, 50e3, 100e3, Inf), labels=FALSE),
         wagegroup=paste0("wage", wagegroup),
         wtdwage=E00200 * weight) %>%
  spread(wagegroup, wtdwage) %>%
  mutate_at(vars(starts_with("wage")), list(~ifelse(is.na(.), 0, .))) %>%
  select(RECID, starts_with("wage"))
ht(wagevals)
# count(wagevals, wagegroup)

constraint_vars <- c("E00100_npos", "E00100_sumval", "E00200_sumval", "E02000_nneg", "wage1", "wage2", "wage3")

ccoef <- ccoef_all %>%
  left_join(wagevals) %>%
  select(constraint_vars)
  
# Get the file sums so we know where we are starting from
(filesums <- calc_constraints(ccoef, constraint_vars))
# (filesums <- calc_constraints(ccoef, constraint_vars, x=1))

# create targets with perturbed values
targ_factors <- list(E00100_npos=1.25,
                     E00100_sumval=0.75,
                     E00200_sumval=.9,
                     E02000_nneg=.8,
                     wage1=1,
                     wage2=.85,
                     wage3=.95) %>% 
  unlist # define multipliers for some values, as a named vector

targets <- filesums[names(targ_factors)] * targ_factors

# Look at our targets
cbind(init_vals=filesums[names(targets)], targets, ratio=targets/filesums[names(targets)])

targets

inputs <- list()
inputs$p <- 2
inputs$weights <- pufsub$weight
inputs$ccoef <- ccoef
inputs$constraint_vars <- constraint_vars
inputs$cc_dense <- c(as.matrix(inputs$ccoef[, inputs$constraint_vars])) # flattens the cc matrix

# create vectors with constraint lower bounds and upper bounds
clb <- targets[inputs$constraint_vars]
cub <- targets[inputs$constraint_vars]

cbind(init_vals=filesums[names(targets)], targets, clb, cub)

eval_jac_g_structure_dense <- define_jac_g_structure_dense(n_constraints=length(inputs$constraint_vars), 
                                                           n_variables=nrow(inputs$ccoef))
# str(eval_jac_g_structure_dense)

eval_h_structure <- lapply(1:length(inputs$weight), function(x) x) # diagonal elements of our Hessian


# set arbitrary bounds for x that fall inside the solution above
xlb <- rep(.3, length(inputs$weights))
xub <- rep(10, length(inputs$weights))

opts <- list("print_level" = 0,
             "file_print_level" = 5, # integer
             "linear_solver" = "ma27", # mumps pardiso ma27 ma57 ma77 ma86 ma97
             "max_iter"=200,
             "output_file" = "prob2_wages.out")
  
result2 <- ipoptr(x0 = rep(1, length(inputs$weights)),
                 lb = xlb,
                 ub = xub,
                 eval_f = eval_f_xtop, # arguments: x, inputs
                 eval_grad_f = eval_grad_f_xtop,
                 eval_g = eval_g_dense, # constraints LHS - a vector of values
                 eval_jac_g = eval_jac_g_dense,
                 eval_jac_g_structure = eval_jac_g_structure_dense,
                 eval_h = eval_h_xtop, # the hessian is essential for this problem
                 eval_h_structure = eval_h_structure,
                 constraint_lb = clb,
                 constraint_ub = cub,
                 opts = opts,
                 inputs = inputs)

# names(result1b) %>% sort
paste0("Status: ", result2$status)
result2$message
paste0("Iterations: ", result2$iterations)
result2$objective
quantile(result2$solution, probs=c(0, .05, .1, .25, .5, .75, .9, .95, 1))

calcsums <- calc_constraints(inputs$ccoef, inputs$constraint_vars, x=result2$solution)
cbind(filesums, targets, clb, calcsums, cub)


```


## Use inequality constraints
```{r optimize3_wages, echo=TRUE}

# create vectors with constraint lower bounds and upper bounds
# define tolerance - for simplicity, use the same for all, but that is not necessary
tol <- .01 # 1% range
clb <- targets[inputs$constraint_vars]
clb <- clb - tol * abs(clb)

cub <- targets[inputs$constraint_vars]
cub <- cub + tol * abs(cub)

# Create a range 

cbind(init_vals=filesums[names(targets)], targets, clb, cub)

eval_jac_g_structure_dense <- define_jac_g_structure_dense(n_constraints=length(inputs$constraint_vars), 
                                                           n_variables=nrow(inputs$ccoef))
# str(eval_jac_g_structure_dense)

eval_h_structure <- lapply(1:length(inputs$weight), function(x) x) # diagonal elements of our Hessian


# set arbitrary bounds for x that fall inside the solution above
xlb <- rep(.3, length(inputs$weights))
xub <- rep(10, length(inputs$weights))

# outfile <- "test2.out"
outfile <- "prob2_wages_inequals.out"

opts <- list("print_level" = 0,
             "file_print_level" = 5, # integer
             "linear_solver" = "ma27", # mumps pardiso ma27 ma57 ma77 ma86 ma97
             "max_iter"=200,
             "output_file" = outfile)
  
result <- ipoptr(x0 = rep(1, length(inputs$weights)),
                 lb = xlb,
                 ub = xub,
                 eval_f = eval_f_xtop, # arguments: x, inputs
                 eval_grad_f = eval_grad_f_xtop,
                 eval_g = eval_g_dense, # constraints LHS - a vector of values
                 eval_jac_g = eval_jac_g_dense,
                 eval_jac_g_structure = eval_jac_g_structure_dense,
                 eval_h = eval_h_xtop, # the hessian is essential for this problem
                 eval_h_structure = eval_h_structure,
                 constraint_lb = clb,
                 constraint_ub = cub,
                 opts = opts,
                 inputs = inputs)

# names(result1b) %>% sort
paste0("Status: ", result$status)
result$message

result$iterations

result$objective
quantile(result$solution, probs=c(0, .05, .1, .25, .5, .75, .9, .95, 1))

calcsums <- calc_constraints(inputs$ccoef, inputs$constraint_vars, x=result$solution)
cbind(filesums, targets, clb, calcsums, cub) %>%
  kable(digits=0, format.args = list(big.mark=","))


```



## Use scaling options
Set objective function scaling (read about this in IPOPT options), as our objective function can be very large. Add the options to the list called opts.

Notice that this takes fewer iterations.

```{r optimize4_wages_scaled, echo=TRUE}

# outfile <- "test2.out"
outfile <- "prob2_wages_inequals_scaled.out"

opts <- list("print_level" = 0,
             "file_print_level" = 5, # integer
             "linear_solver" = "ma27", # mumps pardiso ma27 ma57 ma77 ma86 ma97
             "obj_scaling_factor" = 1e-4, # default 1
             "nlp_scaling_max_gradient" = 100, # default 100
             "max_iter"=200,
             "output_file" = outfile)
  
result <- ipoptr(x0 = rep(1, length(inputs$weights)),
                 lb = xlb,
                 ub = xub,
                 eval_f = eval_f_xtop, # arguments: x, inputs
                 eval_grad_f = eval_grad_f_xtop,
                 eval_g = eval_g_dense, # constraints LHS - a vector of values
                 eval_jac_g = eval_jac_g_dense,
                 eval_jac_g_structure = eval_jac_g_structure_dense,
                 eval_h = eval_h_xtop, # the hessian is essential for this problem
                 eval_h_structure = eval_h_structure,
                 constraint_lb = clb,
                 constraint_ub = cub,
                 opts = opts,
                 inputs = inputs)

paste0("Status: ", result$status)
result$message

result$iterations

result$objective
quantile(result$solution, probs=c(0, .05, .1, .25, .5, .75, .9, .95, 1))


```


  
# TODO: more-complex problems and related complexities

Among other things:

+ How to read IPOPT output
+ How to set the problem up with sparse matrices (if we need to solve HUGE problems)
+ Troubleshooting problems that are hard to solve
+ Running massive problems in parallel (may never need this)



